SNP Analysis

### 2018-07-20 SNP_Prediction_Analysis.Rmd
added cv model build/analysis

```{r setup, include=FALSE}
library(ggplot2)
library(gridExtra)
library(glmnet)
library(pander)
library(readxl)
library(pROC)
library(ggfortify)

@@

load('~/Data/Ellie/Analysis/5e6_arms_all_logR/model_data.Rdata', verbose = T)
load('~/Data/Ellie/Analysis/5e6_arms_all_logR/all.pt.alpha.Rdata', verbose = T)
#alpha = '0'
alpha = '0.9'
fitV = models[[alpha]]
l = performance.at.1se[[alpha]]$lambda

@@

## high CN | ploidy, nonP
highPloidyCN = subset(qcdata, Status == 'NP' & SampleType == 'BE' & (SCA.Ratio >= mean(qcdata$SCA.Ratio,na.rm=T)+ sd(qcdata$SCA.Ratio,na.rm=T)*2 | Ploidy >= 2.3))

@@

@@

```{r, echo=F, fig.height=12, fig.width=6}
grid.arrange(
  ggplot( melt(allDf), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, sWGS'),
  ggplot( melt(arrayDf), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, SNP arrays'),
  ncol=1
)

grid.arrange(
  ggplot(melt(allDf[,1,drop=F]), aes(sample=value)) + stat_qq(),
  ggplot(melt(arrayDf[,1,drop=F]), aes(sample=value)) + stat_qq(),
  ncol=1)

for (i in 1:ncol(arrayDf)) {
  arrayDf[,i] = unit.var(arrayDf[,i], mean(allDf[,i]), sd(allDf[,i]))
}

qqnorm()
sd(allDf[,1])

mean(arrayDf[,1])
sd(arrayDf[,1])

@@

pnr = grep('BLD|gastric', rownames(preds), ignore.case = T)
preds.normal = preds[pnr,]

@@

#offset = log(0.0225)
# preds = preds %>% dplyr::mutate( 
#   'Adj. RR'=RR+offset,  
#   'Adj. Prob'=1/(1+exp(-RR+abs(offset)) )
# )

@@

```{r, warning=F, fig.height=8}
ggplot(preds.normal, aes(Prob)) + geom_histogram(aes(fill=..x..), breaks=seq(0,1,0.1) , show.legend = F) +
  scale_fill_gradientn(colors = myPal,  name='') +
  labs(title='Predictions on BLD/Gastric', y='n Samples', x='Unadjusted Probability')

@@

prog = subset(pred.qc, Status == 'P')
ggplot(prog, aes(Prob)) + facet_grid(~Timepoint) + 
  geom_histogram(aes(fill=..x..), breaks=seq(0,1,0.1) , show.legend = F) +
  scale_fill_gradientn(colors = myPal,  name='') + 
  labs(title='Predictions on progressors', y='n Samples', x='Unadjusted Probability')

nonprog = subset(pred.qc, Status == 'NP')
ggplot(nonprog, aes(Prob)) + facet_grid(~Timepoint) + 
  geom_histogram(aes(fill=..x..), breaks=seq(0,1,0.1) , show.legend = F) +
  scale_fill_gradientn(colors = myPal,  name='') + 
  labs(title='Predictions on non-progressors', y='n Samples', x='Unadjusted Probability')

@@

# Three groups, prog, non prog (max LGD), non prog (max HGD)
pred.qc = pred.qc %>% group_by(Status, PatientID) %>% dplyr::mutate( 
  exclude=(Status == 'NP' & length(which(Pathology == 'HGD')) > 0) )
pred.qc$Status = factor(pred.qc$Status)

## First of all we have samples from patients that probably had a genome duplication, there's no way to measure that in our sWGS data
#subset(pred.qc, `mean(Ploidy)` > 2.7)

@@

```{r echo=F}

pred.qc.ex = subset(pred.qc, !exclude)
cuts = seq(0,1,0.1)

pred.qc.ex$quants = with(pred.qc.ex, cut(Prob, breaks=cuts))
qt = as.data.frame.matrix(table(pred.qc.ex$quants, pred.qc.ex$Status))
qt$quant = rownames(qt)
pander(qt)

get.risk<-function(prob) {
  if (prob < 0.3) return('Low')
  if (prob >= 0.3 & prob < 0.7) return('Moderate')
  if (prob >= 0.7) return('High')
}

@@

roc = pROC::roc(Status ~ Risk, data=pred.qc, auc=T, ci=T, of='thresholds')
roc.plot(roc) + labs(title='All, Risk')

@@

# Exclude the low SCA patients
pred.qc[which(pred.qc$PatientID %in% lowsca$PatientID & pred.qc$EndoID %in% lowsca$EndoID), 'exclude'] = T

per_pt_ex = subset(pred.qc, !exclude) %>% group_by(PatientID, Timepoint, Status) %>% dplyr::summarise(
  max.Prob = max(Prob), max.RR = max(RR), max.Risk = max(Risk)
)
rocppex = pROC::roc(Status ~ max.Prob, data=per_pt_ex, auc=T, ci=T, of='thresholds')
roc.plot(rocppex) + labs(title='Max per patient, Prob')

roct2 = pROC::roc(Status ~ Prob, data=subset(pred.qc, Timepoint != 'T1' & !exclude), auc=T, ci=T, of='thresholds')
roc.plot(roct2) + labs(title='T2, Prob')

@@

ggplot(subset(pred.qc, !exclude), aes(`mean(Purity)`,Prob, color=Risk, shape=Status)) + 
  facet_grid(~Status) + geom_point()

@@

# ggplot(melt(arrayDf), aes(Var2, value)) + geom_point() + 
#   geom_point(data=melt(arrayDf['1006_190V',,drop=F]), color='red') + 
#   geom_point(data=melt(arrayDf['1044_39Z',,drop=F]), color='blue')

# Regions predicted by model
Not evaluated currently
```{r coefs, eval=F}
features = coefs$`0.9`[,1,drop=F]

locs = do.call(rbind.data.frame, c(strsplit(rownames(features), ':|-'), stringsAsFactors=F))
colnames(locs) = c('chr','start','end')
locs[] = lapply(locs[], as.numeric)

feature.locs = makeGRangesFromDataFrame(locs)

gr = makeGRangesFromDataFrame(segments.list$`512`, start.field = 'startpos', end.field = 'endpos', keep.extra.columns = T)

ov = findOverlaps(feature.locs, gr)

feature.locs[unique(queryHits(ov))]

cna = lapply( unique(queryHits(ov)), function(hit) {
  tmp = gr[subjectHits(ov[queryHits(ov) == hit])]
  tmp = subset(as.data.frame(tmp), nMajor+nMinor > 2.3 | nMajor+nMinor < 1.5) %>% group_by(sample, nMajor+nMinor) %>% dplyr::summarise( n=length(sample), medLRR=mean(medLRR), mn.bp=mean(end-start) ) %>% rowwise() %>% 
    dplyr::mutate( type=ifelse(`nMajor + nMinor` >= 3, 1, 0))
})
names(cna) = rownames(features)

fl = sapply(cna, function(x) length(unique(x$sample)))
feature.locs[which(fl == 0)]

cna.count = do.call(rbind, lapply(cna, function(x) table(x$type)))

cna.mx = as.matrix(sapply(cna, function(x) length(unique(x$sample))))
cna.mx = cbind('sample.cnt'=cna.mx, 'loss'=NA,'gain'=NA, 'loss.bp', 'gain.bp')
for (ft in rownames(features)) {
  cna.mx[ft,c('loss','gain')] = c(0,0)
  if (nrow(cna[[ft]]) > 0)
    cna.mx[ft,c('loss','gain')] = c(length(which(cna[[ft]]$type == 0)),length(which(cna[[ft]]$type == 1)))

  median(cna[[ft]] [[cna[[ft]]$type == 0,'mn.bp']] )

}

cna$`1:59820150-64805161`

@@

```{r regression, echo=F, eval=F}
slabels = sample.info[,c('PatientID','Status', 'Samplename')]

missing = setdiff(sample.info$Samplename, rownames(arrayDf))
rows = grep('BLD|gastric', rownames(arrayDf), ignore.case=T)
bld_gastric = arrayDf[rows,]

df = arrayDf[-rows,]

slabels = subset(slabels, Samplename %in% rownames(df))

status = as.integer(as.factor(slabels$Status))-1
names(status) = slabels$Samplename

df = df[-grep(setdiff(rownames(df),names(status)), rownames(df)),]

#df = arrayDf[which(!rownames(arrayDf) %in% c(lowsca$Samplename,highPloidyCN$Samplename)),]

@@

nl = 1000;folds = 10; splits = 5 
sets = create.patient.sets(slabels, folds, splits, 0.2)

#sets = create.patient.sets(subset(sample.info, !Samplename %in% c(lowsca$Samplename,highPloidyCN$Samplename)) [c('PatientID','Samplename','Status')], folds, splits, 0.2)

alpha.values = c(0,0.5,0.9,1)
## ----- All ----- ##
coefs = list(); plots = list(); performance.at.1se = list(); models = list(); cvs = list()
cache.dir = '~/Data/Ellie/Analysis/SNP/'
dir.create(cache.dir, recursive = T, showWarnings = F)
file = paste(cache.dir, 'all.pt.alpha.Rdata', sep='/')
#if (file.exists(file)) {
#  message(paste("loading", file))
#  load(file, verbose=T)
#} else 
  {
  for (a in alpha.values) {
    fit0 <- glmnet(df, status, alpha=a, nlambda=nl, family='binomial', standardize=F)    
    autoplot(fit) + theme(legend.position="none")
    l = fit0$lambda

@@

    performance.at.1se[[as.character(a)]] = subset(cv.patient$lambdas, `lambda-1se`)
    models[[as.character(a)]] = fit0
    cvs[[as.character(a)]] = cv.patient
  }
  save(plots, coefs, performance.at.1se, dysplasia.df, models, cvs, labels, file=file)
  p = do.call(grid.arrange, c(plots[ as.character(alpha.values) ], top='All samples, 10fold, 5 splits'))
  ggsave(paste(cache.dir, '/', 'all_samples_cv.png',sep=''), plot = p, scale = 2, width = 12, height = 10, units = "in", dpi = 300)
}
all.coefs = coefs
save(arrayDf,labels,patient.info, file=paste(cache.dir,'snp.Rdata',sep='/'))
```

>>>>

```{r setup, include=FALSE}
library(ggplot2)
library(gridExtra)
library(glmnet)
library(pander)
library(readxl)
library(pROC)
library(ggfortify)
library(preprocessCore)

@@

load('~/Data/Ellie/Analysis/5e6_arms_all/model_data.Rdata', verbose = T)
load('~/Data/Ellie/Analysis/5e6_arms_all/all.pt.alpha.Rdata', verbose = T)
#alpha = '0'
alpha = '0.9'
fitV = models[[alpha]]
l = performance.at.1se[[alpha]]$lambda
features = coefs[[alpha]]

@@

## high CN | ploidy, nonP
highPloidyCN = subset(qcdata, Status == 'NP' & SampleType == 'BE' & (SCA.Ratio >= mean(qcdata$SCA.Ratio,na.rm=T)+ sd(qcdata$SCA.Ratio,na.rm=T)*2 | Ploidy >= 2.7))

@@

segmentVariance = apply(mergedSegs, 1, var)

@@

```{r, echo=F, fig.height=12, fig.width=6}
## TODO: quantile normalize the SNP arrays by the sWGS values. 
copy = do.call(cbind.data.frame,lapply(1:ncol(arrayDf), function(i) {
  as.data.frame(normalize.quantiles.use.target(arrayDf[,i,drop=F], as.vector(allDf[,i])))
}))

colnames(copy) = colnames(arrayDf)
rownames(copy) = rownames(arrayDf)

grid.arrange(
  ggplot( melt(allDf), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, sWGS'),
  ggplot( melt(arrayDf), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, SNP arrays'),
  ggplot( melt(copy), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, SNP arrays q-normed'),
  ncol=1
)
arrayDf = as.matrix(copy)
```

@@

# offset = log(0.0225)
# preds = preds %>% dplyr::mutate(
#   'Adj. RR'=RR+offset,
#   'Adj. Prob'=1/(1+exp(-RR+abs(offset)) )
# )

pnr = grep('BLD|gastric', preds$Samplename, ignore.case = T)
preds.normal = preds[pnr,]

@@

@@

@@

prog = subset(pred.qc, Status == 'P')
p1 = ggplot(prog, aes(Prob)) + facet_wrap(~Timepoint, scales='free_y',nrow=1) + 
  geom_histogram(aes(fill=..x..), breaks=seq(0,1,0.1) , show.legend = F) +
  scale_fill_gradientn(colors = myPal,  name='') + 
  labs(title='Predictions on progressors', y='n Samples', x='Unadjusted Probability')

nonprog = subset(pred.qc, Status == 'NP')
p2 = ggplot(nonprog, aes(Prob)) + facet_wrap(~Timepoint, scales='free_y', nrow=1) + 
  geom_histogram(aes(fill=..x..), breaks=seq(0,1,0.1) , show.legend = F) +
  scale_fill_gradientn(colors = myPal,  name='') + 
  labs(title='Predictions on non-progressors', y='n Samples', x='Unadjusted Probability')

p3 = ggplot(preds.normal, aes(Prob)) + geom_histogram(aes(fill=..x..), breaks=seq(0,1,0.1) , show.legend = F) +
  scale_fill_gradientn(colors = myPal,  name='') +
  labs(title='Predictions on BLD/Gastric', y='n Samples', x='Unadjusted Probability')

grid.arrange(p1,p2,p3,ncol=1)

@@

# These samples are likely genome duplicated.
dup = subset(pred.qc, `mean(Ploidy)` > 2.7)
cor.test(dup$`mean(Ploidy)`, dup$Prob)
table(dup$Status)

# Three groups, prog, non prog (max LGD), non prog (max HGD)
pred.qc = pred.qc %>% group_by(Status, PatientID) %>% dplyr::mutate( 
  exclude=(Status == 'NP' & (length(which(Pathology == 'HGD')) > 0 | `mean(Ploidy)` > 2.7 )) )
pred.qc$Status = factor(pred.qc$Status)

@@

```{r echo=F}
pred.qc.ex = subset(pred.qc, !exclude)
cuts = seq(0,1,0.1)

pred.qc.ex$quants = with(pred.qc.ex, cut(Prob, breaks=cuts))
qt = as.data.frame.matrix(table(pred.qc.ex$quants, pred.qc.ex$Status))
qt$quant = rownames(qt)
qt$P = round(qt$P/sum(qt$P),2)
qt$NP = round(qt$NP/sum(qt$NP), 2)
pander(qt)

# These are the cutoffs from the sWGS data
get.risk<-function(prob) {
  if (prob < 0.3) return('Low')
  if (prob >= 0.3 & prob < 0.7) return('Moderate')
  if (prob >= 0.7) return('High')
}

@@

table(subset(pred.qc, Risk == 'High' & !exclude)$Status)/table(subset(pred.qc, !exclude)$Status)
table(subset(pred.qc, Risk == 'Moderate' & !exclude)$Status)/table(subset(pred.qc, !exclude)$Status)
table(subset(pred.qc, Risk == 'Low' & !exclude)$Status)/table(subset(pred.qc, !exclude)$Status)

@@

# Exclude the low SCA patients
pred.qc[which(pred.qc$PatientID %in% lowsca$PatientID & pred.qc$EndoID %in% lowsca$EndoID), 'exclude'] = T
rocpp = pROC::roc(Status ~ Prob, data=subset(pred.qc, !exclude), auc=T, ci=T, of='thresholds')
roc.plot(rocpp) + labs(title='Max per patient (excluded low sca/NP tetra), Prob')
pROC::coords(rocpp, 'best')

as.data.frame(subset(pred.qc, Status == 'P' & Prob < 0.5 & !exclude))

# Timepoint 2 only
roct2 = pROC::roc(Status ~ Prob, data=subset(pred.qc, Timepoint != 'T1' & !exclude), auc=T, ci=T, of='thresholds')
roc.plot(roct2) + labs(title='T2, Prob')

per_pt_ex = subset(pred.qc, !exclude) %>% group_by(PatientID, Status) %>% dplyr::summarise(
  max.Prob = max(Prob), max.RR = max(RR), max.Risk = max(Risk)
)
rocppex = pROC::roc(Status ~ max.Prob, data=per_pt_ex, auc=T, ci=T, of='thresholds')
roc.plot(rocppex) + labs(title='Max per patient, Prob')

@@

ggplot(subset(pred.qc, !exclude | ( `mean(SCA.Ratio)` > 0.1 & Status == 'P')) , aes(`mean(Purity)`,Prob, color=Risk, shape=Status)) + facet_grid(~Status) + geom_point()

@@

 ggplot(melt(arrayDf), aes(Var2, value)) + geom_point() + 
  geom_point(data=melt(arrayDf['1006_190V',,drop=F]), color='red') + 
  geom_point(data=melt(arrayDf['1044_39Z',,drop=F]), color='blue')

@@

```{r regression, echo=F, message=F}
slabels = as.data.frame(subset(pred.qc, !exclude, select=c('PatientID','Samplename','Status')))
#slabels = as.data.frame(pred.qc, select=c('PatientID','Samplename','Status'))
rownames(slabels) = slabels$Samplename

df = arrayDf[which(rownames(arrayDf) %in% slabels$Samplename),]

status = as.integer(slabels$Status)-1
names(status) = slabels$Samplename

@@

nl = 1000;folds = 10; splits = 5 
## ----- All ----- ##
coefs = list(); plots = list(); performance.at.1se = list(); models = list(); cvs = list()
cache.dir = '~/Data/Ellie/Analysis/SNP'
dir.create(cache.dir, recursive = T, showWarnings = F)
file = paste(cache.dir, 'all.pt.alpha.Rdata', sep='/')
if (file.exists(file)) {
 message(paste("loading", file))
 load(file, verbose=T)
} else {
  alpha.values = c(0,0.5,0.7,0.9,1)
  sets = create.patient.sets(slabels, folds, splits, 0.2)

  for (a in alpha.values) {
    message(paste('alpha=',a))
    fit0 <- glmnet(df, status, alpha=a, nlambda=nl, family='binomial', standardize=F)    
    autoplot(fit0) + theme(legend.position="none")
    l = fit0$lambda
    if (a > 0) l = more.l(l)

@@

    performance.at.1se[[as.character(a)]] = subset(cv.patient$lambdas, `lambda-1se`)
    models[[as.character(a)]] = fit0
    cvs[[as.character(a)]] = cv.patient
  }
  save(plots, coefs, performance.at.1se, df, models, cvs, labels, file=file)
  p = do.call(grid.arrange, c(plots[ as.character(alpha.values) ], top='All samples, 10fold, 5 splits'))
  ggsave(paste(cache.dir, '/', 'all_samples_cv.png',sep=''), plot = p, scale = 2, width = 12, height = 10, units = "in", dpi = 300)
}
#save(arrayDf,labels,patient.info, file=paste(cache.dir,'snp-models.Rdata',sep='/'))

do.call(grid.arrange, c(plots))

model.performance(performance.at.1se, coefs)$plot + labs(title='SNP CV models')

```{r compare.coefs, echo=F, eval=T}
# Matching
snp.features = coefs$`0.5`

intersect(rownames(snp.features), rownames(features))

cxisft = grep('cx',rownames(snp.features))

rowSums(snp.features[,-1])/50

snp.features = do.call(rbind.data.frame, c(strsplit(rownames(snp.features)[-cxisft], ':|-'), stringsAsFactors=F))
colnames(snp.features) = c('chr','start','end')

swgs.features = do.call(rbind.data.frame, c(strsplit(rownames(features), ':|-'), stringsAsFactors=F))
colnames(swgs.features) = c('chr','start','end')

snp.features = makeGRangesFromDataFrame(snp.features)
snp.features = snp.features[-which(width(snp.features) > 5e6)] # No arms

swgs.features = makeGRangesFromDataFrame(swgs.features)

dist = distanceToNearest(swgs.features,snp.features)

swgs.features[queryHits(dist)]
snp.features[subjectHits(dist)]

match = dist[elementMetadata(dist)$distance != 0]
elementMetadata(dist)$bins = round(elementMetadata(dist)$distance/5e6)

x1 = dist[elementMetadata(dist)$bins <= 2]

swgs.features[queryHits(x1)]
snp.features[subjectHits(x1)]
```

```{r loo}

slabels$Prediction = NA
slabels$RR = NA

file = paste(cache.dir, 'loo.Rdata', sep='/')
if (file.exists(file)) {
  load(file, verbose=T)
} else {
  select.alpha = 1
  secf = coefs[[select.alpha]]
  a = select.alpha

  performance.at.1se = c(); plots = list(); fits = list(); nzcoefs = list()
  # Remove each patient (LOO)
  for (pt in unique(slabels$PatientID)) {
    print(pt)
    samples = subset(slabels, PatientID != pt)$Samplename

    train.rows = which(rownames(df) %in% samples)
    training = df[train.rows,]
    test = as.matrix(df[-train.rows,])
    if ( nrow(test) == ncol(df) ) test = t(test)

    # Predict function giving me difficulty when I have only a single sample, this ensures the dimensions are the same
    sparsed_test_data <- Matrix(data=0, nrow=nrow(test),  ncol=ncol(training),
                                dimnames=list(rownames(test),colnames(training)), sparse=T)
    for(i in colnames(arrayDf)) sparsed_test_data[,i] = test[,i]

    # Fit generated on all samples, including HGD
    fitLOO <- glmnet(training, status[train.rows], alpha=a, family='binomial', nlambda=nl, standardize=F) # all patients
    l = fitLOO$lambda

    cv = crossvalidate.by.patient(x=training, y=status[train.rows], lambda=l, a=a, nfolds=folds, splits=splits,
                                  pts=subset(sets, Samplename %in% samples), fit=fitLOO, standardize=F)

    plots[[pt]] = arrangeGrob(cv$plot+ggtitle('Classification'), cv$deviance.plot+ggtitle('Binomial Deviance'), top=pt, ncol=2)
    fits[[pt]] = cv

    if ( length(cv$lambda.1se) > 0 ) {
      performance.at.1se = c(performance.at.1se, subset(cv$lambdas, lambda == cv$lambda.1se)$mean)

      coef.1se = coef(fitLOO, cv$lambda.1se)[rownames(secf),]

      nzcoefs[[pt]] = as.data.frame(non.zero.coef(fitLOO, cv$lambda.1se))

      #coefs[[pt]] = coef(fitLOO, cv$lambda.1se)[rownames(secf),]
      #coef.stability(coef.1se, cv$non.zero.cf)

      logit <- function(p){log(p/(1-p))}
      inverse.logit <- function(or){1/(1 + exp(-or))}

      pm = predict(fitLOO, newx=sparsed_test_data, s=cv$lambda.1se, type='response')
      or = predict(fitLOO, newx=sparsed_test_data, s=cv$lambda.1se, type='link')

      slabels[which(slabels$Samplename %in% rownames(pm)),'Prediction'] = pm[,1]
      slabels[which(slabels$Samplename %in% rownames(pm)),'RR'] = or[,1]

    } else {
      warning(paste(pt, "did not have a 1se"))
    }
  }
  save(plots, performance.at.1se, nzcoefs, fits, slabels, file=file)
}

snp.roc = pROC::roc(Status~Prediction,slabels, auc=T, ci=T, of='thresholds')
roc.plot(snp.roc, title="SNP LOO Model ROC")

ggplot(cbind.data.frame('performance'=performance.at.1se, 'type'='snp'), aes(y=performance, x=type, group=type,fill=type)) + 
  geom_boxplot(outlier.colour = NA, show.legend = F) + geom_jitter(width=0.2) + geom_label(data=cbind.data.frame('performance'=range(performance.at.1se), 'type'='snp'), aes(label=round(performance,2)), nudge_x=0.1) + theme_bw() + theme(legend.position = 'none') + labs(x='', y='Classification', title='LOO Performance SNP')

file = paste(cache.dir, 'all.pt.alpha.Rdata', sep='/')
load(file, verbose=T)

fit.snp = models$`1`

probs = predict(fit.snp, newx = dysplasia.df, s = performance.at.1se$`1`$lambda,type='response' )

swgs.pred.snp = cbind.data.frame('Prob'=probs[,1], 'Status'=swgs_labels[rownames(probs)])
ssroc = pROC::roc(Status~Prob, data=swgs.pred.snp, ci=T, auc=T, of='thresholds')
roc.plot(ssroc, title='sWGS prediction on SNP model')
```
